---
title: How does early-stage funding affect the long-term success or failure of startups?
subtitle: Data Analysis Project
author: Joshua Wong
format: 
  html:
    self-contained: true
    code-tools: true
    toc: true
    code-fold: true
editor: source
execute: 
  error: true
  echo: true
  message: false
  warning: false
---

## Introduction

As you've heard in the news and social media, startups are becoming increasingly popular in public interest. According to [Forbes](https://www.forbes.com/advisor/business/what-is-a-startup/), a startup is defined as "a businesses that wants to disrupt industries and change the world---and do it all at scale." Oftentimes, it's high-risk high-reward, especially the earlier you are in it. This crucial early stage shapes the outcome of the business, and to keep it alive with high-growth, many founders raise rounds to fund their company. This paper aims to analyze the impact of early-stage funding on the longer-term success (or failure) of startups. For this research question, we will define "early-stage funding" as a startup that's raised an angel investment and/or a series A investment. For more information about startup rounds, you may find some useful information [here](https://www.forbes.com/sites/alejandrocremades/2018/12/26/how-funding-rounds-work-for-startups/?sh=49cbfb767386). The "success" of a startup will be defined in two separate ways: if the startup was in the "Top 500" startups (according to the data) and if the startup was acquired. As defined, please note that both success metrics will have their own respective analyses and findings, and correlation between the success metrics will not be performed in the scope of this article.

### Data Description

The dataset used in this analysis, [startup-data.csv](https://www.kaggle.com/datasets/manishkc06/startup-success-prediction/data), includes comprehensive information regarding various startups including their funding history and current status. This dataset has data about different funding rounds as well as geographical information, average number of funding participants, and the current status of each startup (e.g., acquired, closed). The dataset provides a snapshot of 923 startups betweeen 1984 and 2013.

### Objective

The objective of this data analysis is to examine the relationship between early-stage funding and the long-term success of startups. A key hypothesis is that startups receiving early-stage funding have a higher likelihood of long-term success compared to those without. This analysis aims to validate / challenge this hypothesis through some data exploration and statistical testing of the dataset.

### Approach

-   Identify and categorize startups based on their early-stage funding status (VC, angel investment, and Round A funding).
-   Analyze the correlation between early-stage funding and the long-term success or failure of these startups.
-   Perform statistical tests to ascertain the strength and significance of these correlations.
-   Use visualizations to illustrate key findings and trends in the data.

# Setup

### Packages

```{r setup}
# Load necessary packages
library(tidyverse)
library(dplyr)
library(ggplot2)
library(readr)
library(here)
library(lubridate)
library(DT)
library(scales)
```

### Load Data

```{r load}
startup_data <- read_csv(here("data/startup-data.csv"))
```

### Interactive Preview of Cleaned Data

The original dataset underwent several cleaning processes:

-   Conversion of relevant columns to appropriate data types.
-   Categorization of startups based on their funding status (VC, angel investment, and Round A, etc.).
-   Filtering to focus on critical attributes like funding status, geographic location, and current status.

```{r cleaning}

# Remove unnecessary columns
startup_data <- startup_data |>
  select(-c('Unnamed: 0', 'Unnamed: 6'))

# Convert data types
startup_data <- startup_data |>
  mutate(
    latitude = as.numeric(latitude),
    longitude = as.numeric(longitude),
    zip_code = as.character(zip_code),
    founded_at = mdy(founded_at),
    closed_at = mdy(closed_at),
    first_funding_at = mdy(first_funding_at),
    last_funding_at = mdy(last_funding_at),
    across(starts_with('age'), as.numeric),
    funding_total_usd = as.numeric(funding_total_usd),
    avg_participants = as.numeric(avg_participants),
    across(contains('is_'), as.logical),
    status = as.factor(status)
  )

# Drop NA for status column
startup_data <- startup_data |>
  drop_na(status)

# Create a new column 'early_stage_funding' to indicate if a startup has received any early-stage funding
startup_data <- startup_data |>
  mutate(early_stage_funding = has_angel | has_roundA)
```

## Interactive Preview of Cleaned Data

```{r data-preview}
datatable(startup_data, options = list(pageLength = 5, scrollX = TRUE))
```

#### Looking at Date Ranges to Determine Bounds of Data Snapshot

```{r date-ranges}
check_dates <- startup_data |>
  mutate(
    founded_at = as.Date(founded_at, format = "%m/%d/%Y"),
    first_funding_at = as.Date(first_funding_at, format = "%m/%d/%Y"),
    last_funding_at = as.Date(last_funding_at, format = "%m/%d/%Y")
  )

check_dates |>
  summarise(
    min_founded_at = min(founded_at, na.rm = TRUE),
    max_founded_at = max(founded_at, na.rm = TRUE),
    min_first_funding_at = min(first_funding_at, na.rm = TRUE),
    max_first_funding_at = max(first_funding_at, na.rm = TRUE),
    min_last_funding_at = min(last_funding_at, na.rm = TRUE),
    max_last_funding_at = max(last_funding_at, na.rm = TRUE)
  )
```

#  Analysis

```{r top-500-percentage}

startup_data$early_stage_funding <- factor(if_else(startup_data$has_angel == 1 | startup_data$has_roundA == 1, "Yes", "No"), 
                                           levels = c('No', 'Yes'))

startup_data$is_top500_factor <- factor(startup_data$is_top500, 
                                        levels = c(TRUE, FALSE), 
                                        labels = c("Top 500", "Not Top 500"))

count_rates <- startup_data |>
  count(early_stage_funding, is_top500_factor) |>
  group_by(early_stage_funding) |>
  mutate(prop = n / sum(n)) |>
  ungroup()

ggplot(count_rates, aes(x = early_stage_funding, y = prop, fill = is_top500_factor)) +
  geom_bar(stat = "identity") +
  scale_y_continuous(labels = percent_format()) +
  scale_fill_manual(values = c("Top 500" = "palegreen", "Not Top 500" = "lightcoral")) +
  labs(
    title = "Proportion of Top 500 Startups by Early-Stage Funding Status",
    x = "Received Early-Stage Funding?",
    y = "",
    fill = "Top 500 Status"
  ) +
  theme_minimal() +
  geom_text(aes(label = scales::percent(prop, accuracy = 1)), 
            position = position_stack(vjust = 0.5), 
            size = 3, 
            color = "black")
```

```{r acquired-percentage}

startup_data$early_stage_funding <- factor(if_else(startup_data$has_angel == 1 | startup_data$has_roundA == 1, "Yes", "No"), 
                                           levels = c('No', 'Yes'))

startup_data$status_factor <- factor(startup_data$status, levels = c('acquired', 'closed'))

count_rates <- startup_data |>
  count(early_stage_funding, status_factor) |>
  group_by(early_stage_funding) |>
  mutate(prop = n / sum(n)) |>
  ungroup()

ggplot(count_rates, aes(x = early_stage_funding, y = prop, fill = status_factor)) +
  geom_bar(stat = "identity") +
  scale_y_continuous(labels = percent_format()) +
  scale_fill_manual(values = c("acquired" = "palegreen", "closed" = "lightcoral")) +
  labs(
    title = "Proportion of Startups by Status and Early-Stage Funding",
    x = "Received Early-Stage Funding?",
    y = "",
    fill = "Startup Status"
  ) +
  theme_minimal() +
  geom_text(aes(label = scales::percent(prop, accuracy = 1)), 
            position = position_stack(vjust = 0.5), 
            size = 3, 
            color = "black")
```

Looking at the two charts above, we can see that the startups that received early stage funding hold a majority (positive relation) with both of our success metrics.\
\
Regression

I'm using `glm()` because it is more fitting for our data than `lm()`, where we have binary outcome variables. With `glm()` we can use the "binomial" option. You may read more about it [here](https://sscc.wisc.edu/sscc/pubs/glm-r/#glm-families).

### Regression Function

```{r regression-function}
run_logistic_regression <- function(data, response, predictor) {
  formula <- as.formula(paste(response, "~", predictor))
  model <- glm(formula, data = data, family = "binomial", maxit = 1000)
  return(summary(model))
}
```

#### Example Usage

```{r func-example-usage}
startup_data$status_factor <- ifelse(startup_data$status == "acquired", 1, 0)

logistic_model_angel <- run_logistic_regression(startup_data, "status_factor", "has_angel")

logistic_model_series_a <- run_logistic_regression(startup_data, "status_factor", "has_roundA")

print(logistic_model_angel)
print(logistic_model_series_a)
```

### 

## Early Stage Funding Regression with "Success" Metrics

### Success Metric - Top 500

```{r early-stage-regression}
startup_data$is_top500_factor <- as.numeric(startup_data$is_top500)

startup_data$early_stage_funding <- ifelse(startup_data$early_stage_funding == "Yes", 1, 0)
run_logistic_regression(startup_data, "is_top500_factor", "early_stage_funding")
```

In this logistic regression analysis, we can see that the coefficient for `early_stage_funding` is negative, suggesting that early-stage funding may decrease the likelihood of being in the top 500 startups compared to not. The results are statistically significant, with a p-value of 0.0206. However the effect size is relatively small and one should note that many other factors play a role in affecting startup success.

### Success Metric - Acquired

```{r acquired-regression}
run_logistic_regression(startup_data, "status_factor", "early_stage_funding")
```

In this logistic regression analysis, we can see that the coefficient for `early_stage_funding` is positive, suggesting that early-stage funding may increase the likelihood of a startup being acquired. However the results are not statistically significant, with a p-value of 0.126.

# Conclusion

In this analysis, we graphed the relation between the proportion of startups who raised early funding rounds with both our success metrics of being in the top 500 and being acquired. From those results, we saw that there was a positive relationship and that there may be some indication of it being significant. After performing a logistic regression on both success metrics, the results suggest that there is no overwhelmingly statistically significant effect of early-stage funding on startup success. The findings indicate that other factors likely play a more substantial role in determining long-term success or failure of startups.
